---
title: "Machine Learning Course Project"
author: "OLAJIDE ADEMOLA"
date: "10/12/2020"
output: html_document
---



## Project Description
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.


## Aim
The goal of this project is to predict the manner in which they did the exercise. This is the “classe” variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. use your prediction model to predict 20 different test cases.

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis.

2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.

## Code and Results

# Packages, Libraries, Seed

Installing packages, loading libraries, and setting the seed for reproduceability:
```{r , echo=TRUE}
library(caret)
#install.packages("caret")
#install.packages("randomForest")
#install.packages("rpart")
## Loading required package: lattice
## Loading required package: ggplot2
```

```{r , echo=TRUE}
library(randomForest) #Random forest for classification and regression
## randomForest 4.6-10
## Type rfNews() to see new features/changes/bug fixes.
```


```{r , echo=TRUE}
library(rpart) 
# Regressive Partitioning and Regression trees
library(rpart.plot) 
# Decision Tree plot
# setting the overall seed for reproduceability
set.seed(1234)
```

## Loading data sets and preliminary cleaning

First we want to load the data sets into R and make sure that missing values are coded correctly.
Irrelevant variables will be deleted.
Results will be hidden from the report for clarity and space considerations.
```{r , echo=TRUE}


# data load and clean up
trainingset <- read.csv("pml-training.csv",  na.strings=c("NA","#DIV/0!", ""))
testingset <- read.csv("pml-testing.csv",  na.strings=c("NA","#DIV/0!", ""))

# Perform exploratory analysis - 
# dim(trainingset); dim(testingset); #summary(trainingset); summary(testingset); #str(trainingset); str(testingset); head(trainingset); #head(testingset);               

# Delete columns with all missing values
trainingset<-trainingset[,colSums(is.na(trainingset)) == 0]
testingset <-testingset[,colSums(is.na(testingset)) == 0]

# Delete variables are irrelevant to our current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). 
trainingset   <-trainingset[,-c(1:7)]
testingset <-testingset[,-c(1:7)]

# partition the data so that 75% of the training dataset into training and the remaining 25% to testing
dim(trainingset)
dim(testingset)
head(trainingset)
head(testingset)


traintrainset <- createDataPartition(y=trainingset$classe, p=0.75, list=FALSE)
TrainTrainingSet <- trainingset[traintrainset, ] 
TestTrainingSet <- trainingset[-traintrainset, ]

dim(TrainTrainingSet)
dim(TestTrainingSet)
head(TrainTrainingSet)
head(TestTrainingSet)
# The variable "classe" contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the TrainTrainingSet data set and # compare one another.


plot(TrainTrainingSet$classe, na.rm = TRUE), col="yellow", main="Plot of levels of variable classe within the TrainTrainingSet data set", xlab="classe", ylab ="Frequency")
```

From the graph above, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent with more than 4000 occurrences while level D is the least frequent with about 2500 occurrences.

# First prediction model: Using Decision Tree

```{r , echo=TRUE}
model1 <- rpart(classe ~ ., data=TrainTrainingSet, method="class")

prediction1 <- predict(model1, TestTrainingSet, type = "class")

# Plot the Decision Tree
rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```


```{r , echo=TRUE}
# Test results on our TestTrainingSet data set:
confusionMatrix(prediction1, TestTrainingSet$classe)
```
Second prediction model: Using Random Forest

```{r , echo=TRUE}
model2 <- randomForest(classe ~. , data=TrainTrainingSet, method="class")

# Predicting:
prediction2 <- predict(model2, TestTrainingSet, type = "class")

# Test results on TestTrainingSet data set:
confusionMatrix(prediction2, TestTrainingSet$classe)
```

# Decision

As expected, Random Forest algorithm performed better than Decision Trees.
Accuracy for Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to 0.739 (95% CI: (0.727, 0.752)) for Decision Tree model. The random Forest model is choosen. The accuracy of the model is 0.995. The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

## Submission
```{r , echo=TRUE}
# predict outcome levels on the original Testing data set using Random Forest algorithm
predictfinal <- predict(model2, testingset, type="class")
predictfinal
```
## References
[1] Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human ’13) . Stuttgart, Germany: ACM SIGCHI, 2013.

[2] Krzysztof Gra??bczewski and Norbert Jankowski. Feature Selection with Decision Tree Criterion.

<https://rstudio-pubs-static.s3.amazonaws.com/52063_0220dab7789142228e34d18ec0df6f3e.html>